---
title: "HLM Preprocessing"
author: "Paul Sylvia"
date: "2025-05-21"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Preparing the Level 1 Dataset

The level 1 dataset is a representative of nearly all US counties. The initial step is to load the necessary libraries, set the working directory and load the necessary datasets. We load data taken from the US census and County Health Rankings and Roadmaps initiative, as well as AHA hospital data to ensure only the involvement of level 1 counties clustered later-on.

```{r, warning=FALSE, message=FALSE}
library.list <- c("dplyr", "psych", "mice")

for (i in 1:length(library.list)) {
  if (!library.list[i] %in% rownames(installed.packages())) {
    install.packages(library.list[i])
  }
  library(library.list[i], character.only = TRUE)}

county_dat <- read.csv('county_MH.csv') ##Primary county-level dataset
social_capital <- read.csv('social_capital.csv') ##Dataset for the social capital variable

#Merge the county data into one dataset, each county is one case
social_capital$X5.digit.FIPS.Code <- social_capital$FIPS.Code
county_dat <- county_dat %>%
  right_join(social_capital, by = "X5.digit.FIPS.Code")

#Add an HRR ID variable to the dataset
HRR_county_crosswalk <- read.csv('HRR_county_crosswalk.csv') ##this dataset tells which HRR a county belongs to
county_dat$HRR <- HRR_county_crosswalk$hrr[match(county_dat$X5.digit.FIPS.Code, HRR_county_crosswalk$county)] #adds HRR variable to L1 dataset
```

The variables of interest have messy names. Here we are renaming them and performing some rescalings.

```{r}
 #Cleaning variable names
county_dat$social_capital <- county_dat$Putnam
county_dat$poverty_rate <- county_dat$Poverty.rate.x/100 #transform to a better scale for HLM
county_dat$population <- county_dat$Population.raw.value
county_dat$suicide <- county_dat$Crude.suicide.rate
county_dat$overdose <- county_dat$overdose/county_dat$population
county_dat$rural <- county_dat$X..Rural.raw.value
county_dat$med.inc <- county_dat$Median.household.income
county_dat$black <- county_dat$X..Non.Hispanic.Black.raw.value
county_dat$hispanic <- county_dat$X..Hispanic.raw.value
county_dat$older <- county_dat$X..65.and.Older.raw.value
county_dat$younger <- county_dat$X..Below.18.Years.of.Age.raw.value
county_dat$unemployment <- county_dat$Unemployment.raw.value
county_dat$med.inc <- county_dat$Median.Household.Income.raw.value
county_dat$gini <- (county_dat$gini-min(county_dat$gini, na.rm=TRUE))/(max(county_dat$gini, na.rm=TRUE)-min(county_dat$gini, na.rm=TRUE)) ##rescales so the smallest value in the dataset==0 and the largest==1. This is done so a one unit change in 'gini' is interpretable

```

Finally, we declare the level-1 dataset and filter out certain cases with too much missing data, then we impute the remaining missing data with predictive mean matching.

```{r, warning=FALSE, message=FALSE}
L1 <- county_dat %>%
  select(HRR, X5.digit.FIPS.Code, overdose, hispanic, older, younger, population, 
         black, rural, gini, med.inc, employment, social_capital, poverty_rate) %>%
  filter(is.na(HRR)==FALSE) #removes cases with no crosswalk

print(paste("Total Missing Data: ", round(sum(is.na(L1))/(nrow(L1)*ncol(L1)),digits=3), sep=""))
L1_listwise <-L1[is.na(L1$employment)==FALSE & is.na(L1$med.inc)==FALSE,]

print(paste("Ammount of missing data imputed: ", round(sum(is.na(L1))/(nrow(L1)*ncol(L1)),digits=3), sep=""))
imp <- mice(L1_listwise[,c(2:9,12)], method="pmm", m = 1, maxit = 1, seed = 123)##predictive mean matching

L1_imp <- cbind(L1_listwise[,c(1,10:11,13)], county_code = L1_listwise$X5.digit.FIPS.Code, complete(imp))
print(paste("Missing data after imputation: ", round(sum(is.na(L1_imp))/(nrow(L1)*ncol(L1))), sep=""))
head(L1_imp)

```

This next step guarantees the cases all have a cluster match.

```{r}
hospital_data <- read.csv('hospital_data_23.csv') ##Primary HRR-level dataset
hospital_data$HRR <- HRR_county_crosswalk$hrr[match(hospital_data$FCOUNTY, HRR_county_crosswalk$county)] #adds HRR variable to L2 dataset

temp <- unique(L1_imp$HRR)[is.na(match(unique(L1_imp$HRR), unique(hospital_data$HRR)))] #identify cluster non-matches
L1_imp <- L1_imp[-which(L1_imp$HRR %in% temp),] ##Remove cases with no cluster matches in L2 dataset
print(paste("Proportion of Cases w/out Cluster Matches: ", round(1-nrow(L1_imp)/nrow(county_dat), digits=3), sep = "")) ##Proportion of cases removed

```

Rescaling Before Aggregating and Centering

```{r}
L1_imp$med.inc <- as.numeric(scale(L1_imp$med.inc))
L1_imp$cluster_population <- L1_imp$population##for scaling Hospital data later
L1_imp$population <- as.numeric(scale(L1_imp$population))
names(L1_imp)
#Grand Mean centering
L1_imp_grand <- as.data.frame(lapply(L1_imp[,-c(1,5,15)], function(x) x - mean(x)))
L1_imp_grand <- cbind(L1_imp[,c(1,5,15)], L1_imp_grand)
names(L1_imp_grand)
colnames(L1_imp_grand)[-(1:3)] <- paste0(colnames(L1_imp_grand)[-(1:3)], "_grand")

write.csv(file = "L1_dataset.csv", L1_imp_grand, row.names=FALSE)
```
